---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# s3 

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/s3)](https://CRAN.R-project.org/package=s3)
[![R-CMD-check](https://github.com/geomarker-io/s3/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/geomarker-io/s3/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

s3 is an R package designed to download files hosted on AWS S3.

Downloads can be saved to a location shared across R sessions/users and download functions always return paths to downloaded files. Files already present in the download location will be used before trying to download a file again. This means more concise code for reading/loading files after downloading them (if they are not already downloaded). For example:

```{r, eval = FALSE}
s3::s3_get("s3://modis-aod-nasa/2020.05.22.tif")
```

## Installation

Install the CRAN latest release inside `R` with:

```r
install.packages("s3")
```

Install the development version from GitHub:

```r
# install.packages("remotes")
remotes::install_github("geomarker-io/s3")
```

## Usage

### Downloading Files

```{r}
library(s3)
```

Download a single file specified by its [URI](#URI) with:

```{r, include=FALSE}
Sys.unsetenv(c('AWS_ACCESS_KEY_ID', 'AWS_SECRET_ACCESS_KEY'))
```

```{r}
s3_get("s3://geomarker/testing_downloads/mtcars.rds")
```

If a file has already been downloaded or already exists, then it will not be re-downloaded:

```{r}
s3_get("s3://geomarker/testing_downloads/mtcars.rds")
```

Download multiple files with:

```{r}
s3_get_files(c(
          "s3://geomarker/testing_downloads/mtcars.rds",
          "s3://geomarker/testing_downloads/mtcars_again.rds"
        ),
	confirm = FALSE)
```

### Private Files

Downloading private files requires the name of the S3 bucket's region (this is determined automatically when the file is public):

```{r, eval = FALSE}
s3_get("s3://geomarker/testing_downloads/mtcars_private.rds", region = "us-east-2")
```

#### Setting up AWS credentials

You must have the appropriate AWS S3 credentials set to gain access to non-public files. As with other AWS command line tools and R packages, you can use the environment variables `AWS_ACCESS_KEY_ID` and `AWS_SECRET_ACCESS_KEY` to gain access to such files. 

It is highly recommended to setup your environment variables outside of your R script to avoid including sensitive information within your R script. This can be done by exporting environment variables before starting R (see [AWS CLI documentation](https://docs.aws.amazon.com/cli/latest/userguide/cli-configure-envvars.html) on this) or by defining them in a `.Renviron` file (see `?.Renviron` within `R`).

You can use the internal helper function to check if AWS key environment variables are set.

```{r}
s3:::check_for_aws_env_vars()
```

### Downloaded file paths

Files are saved within a directory structure matching that of the S3 URI; this directory is created if necessary. `s3_get` and `s3_get_files` both invisibly return the file path(s) of the downloaded files so that they can be further used to access the downloaded files. This makes it possible for different users with different operating systems and/or different project file structures and locations to utilize a downloaded S3 file without changing their source code:

```{r}
s3_get("s3://geomarker/testing_downloads/mtcars.rds") |>
    readRDS()
```

### Customizing download location

By default, files downloaded from S3 will be stored in a folder called `s3_downloads` located within the current working directory. This can be changed when downloading files by using the `download_folder` argument:

```{r}
s3_get("s3://geomarker/testing_downloads/mtcars.rds",
       download_folder = fs::path_home('~/Desktop/s3_downloads'))
```

This can also be changed for the entire session by using the option `s3.download_folder`. For example, specifying `options(s3.download_folder = /scratch/broeg1/s3_downloads)` will write all downloaded files to `/scratch/broeg1/s3_downloads`.

Using a folder for all downloads will prevent duplication of files within different working directories, instead allowing all R sessions to access these files. This could be combined with something like [`rappdirs`](https://github.com/r-lib/rappdirs) to share files across users or temporarily cache them.

As above, this feature also allows different users to store downloaded S3 files in different locations (e.g. a network mounted drive, a scratch folder on a high performance cluster, an external hard drive, a temporary directory) without having to change their R script to specify file paths specific to their computer.

### URI

URI stands for [Universal Resource Identifier](https://en.wikipedia.org/wiki/Uniform_Resource_Identifier), which is a federated and extensible naming system. In practice, this means that a URI is a character string that can uniquely identify a particular resource. An example of a commonly used URI scheme is `https` for identifying web resources, e.g. `https://r-project.org`.

Here, we use the `s3` scheme as defined by AWS. For example, the URI for a file hosted on S3 called `mtcars.rds` in a bucket called `geomarker` in a folder called `test_downloads` would be:

 `s3://geomarker/test_downloads/mtcars.rds`

```{r, include = FALSE}
fs::dir_delete(fs::path_wd('s3_downloads'))
fs::dir_delete(fs::path_home('~/Desktop/s3_downloads'))
```

